---
title: "Trend in Chlorophyll"
author: "Alene Onion"
date: "Feb, 2020"
output:  
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## All Temp Plots

These lakes have at least one data point in each of the past 4 decades.
With the naked eye, you can see that the majority of these waterbodies have declining slopes.

It's also important to note that very few oligotrophic lakes are included in this data set. I believe this is because the focus of the program in the 80s and 90s was to address water quality problems. This was the similar focus nationally via the Clean Water Act. Unfortunately, this means it's difficult to track trends in NYS oligotrophic waters. We have modified our sampling strategy to better capture a balance of conditions moving forward.

```{r, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
temp<-data
#load packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(RColorBrewer)
library(lubridate)
#set date as date
temp$SAMPLE_DATE<-as.Date(temp$SAMPLE_DATE,"%m/%d/%Y")

#pull temp samples
temp<-temp %>% 
  filter(Characteristic.Name %in% c('CHLOROPHYLL A')) %>% 
  select(WATER,LAKE_ID,SAMPLE_DATE,Characteristic.Name,Result.Value,Result.Unit,INFO_TYPE,DATA_PROVIDER) %>% 
  distinct(temp)

#pull year
temp<-temp %>% 
  mutate(year = format(SAMPLE_DATE, "%Y")) %>%
  mutate(year = as.numeric(year)) %>% 
  mutate(decade = year - year %% 10)
temp<-distinct(temp)
temp$Result.Value<-as.numeric(temp$Result.Value)

#pull waterbodies with at least one sample in each decade
decade<-unique(temp[c('LAKE_ID','decade')])
decade<-decade %>% 
  group_by(LAKE_ID) %>% 
  summarize(n()) %>% 
  ungroup()
decade$decades<-decade$`n()`
decade$`n()`<-NULL
decade<-decade[decade$decades>3,]
temp<-merge(decade,temp,by=c('LAKE_ID'),all.x = TRUE)
rm(decade)
temp$decade<-NULL
temp$decades<-NULL

####################################################################################
#subset to just the routine sites
slopes<-read.csv("trend.sites.slopes.provider.csv")
temp<-merge(temp,slopes,all.x = TRUE)
temp<- temp %>% 
#Routine
#    filter(LAKE_ID %in% c('1004PLA0254','1102BAB1109','1310QUE0057','0801SEC0782B','1104SCH0374','0906BON0024','1302WAC0117','1005GLE0441','1404OQU0383','0201CUB0115','0602MOR0152','0602MEL0039','1401ANA0251','0202CHA0122','0202FIN0153','1402WOL0037','1501LUC0982B','1301ORA0340','1402GUY0014','1004AUG0213','0906BUT0054','0705DUC0222','0302SOD0096','1311DUA0276A','1201GAL0563','1701LFR0662','1701CAN0889')) %>% 
#Index
#  filter(LAKE_ID %in% c('1004PLA0254','1102BAB1109','1101BAL1090','0202CHA0122','0201CUB0115','1311DUA0276A','1104EAG0438','1005GLE0441','0104JAV0152','1401ANA0251','0906BON0024','0602MOR0152','0602PET0078','1302WAC0117','1701LFR0662','1701LLO0708','0404LOO0079','0602MEL0039','0906MIL0055','1004MIR0250','1404OQU0383','1310QUE0057','1104SCH0374','0502SMI0066')) %>% 
#CSLAP
#  filter(DATA_PROVIDER %in% c('CSL')) %>% 
  arrange(TSI.slope,LAKE_ID,SAMPLE_DATE)

    temp$lake<-paste(temp$WATER," (",temp$LAKE_ID,")",sep="")
    temp$day<-yday(temp$SAMPLE_DATE)
    params<-unique(temp$lake)
    nparams<-length(params)
    #p<-list()

    for(j in 1:nparams){
      display<-paste("  \n","###",params[j],"  \n",sep="")
      cat(display)
      rm(display)
      
      temp1<-temp[temp$lake==params[j],]
      temp1<-temp1[!is.na(temp1$Characteristic.Name),]
      library(dplyr)

      if((length(temp1$Result.Value))!=0){
        yaxistitle<-unique(temp1$Result.Unit)
          print(p<- ggplot(temp1, aes(x = day, y = Result.Value,color = year)) +
                  theme_minimal()+
                  geom_point(size=2,stroke=2)+
                  labs(y=yaxistitle,x="Day of Year",colour = ""))
      }
      rm(list=c('temp1','testinfo','yaxistitle'))
      cat("\n\n")
    }
    
    cat("  \n")
    cat("  \n")
  rm(list=c('temp','j','params','nparams','p','display'))


```
